Regression models for Cylindrical data in Psychology

### 2019-01-29 Manuscript/ManuscriptMarkdownCV.tex
Update manuscript

\authornote{Add complete departmental affiliations for each
author here. Each new line herein must be indented, like this line.

Enter author note here.

Correspondence concerning this article should be addressed to Jolien
Cremers, Postal address. E-mail:
\href{mailto:joliencremers@gmail.com}{\nolinkurl{joliencremers@gmail.com}}}

@@

In social sciences the use of cylindrical data is very common. Such data
consist of a linear and a circular component. Gurtman (2011) refers to
such data as vectors, with a directional measure (i.e., the circular
component) and a measure indicating the magnitude (i.e., linear
component). Many established models in psychology are often referred to
as circular or circumplex models, but those models are cylindrical.
Examples of such cylindrical models are the interpersonal
circle/circumplex (Leary, 1957; Wiggins, 1996; Wubbels, Brekelmans,
Brok, \& Tartwijk, 2006), the circumplex of affect (Russell, 1980), the
circumplex of human emotion (Plutchik, 1997) or the model of human
values (Schwartz, 1992). \newline \indent Also, many of the more recent
types of data that are studied in psychology are cylindrical. For
example, research on human navigation uses data where distance (i.e.,
linear component) and direction (i.e., the linear component) are of
interest (Chrastil \& Warren, 2017) or in eye-tracking, the saccade data
also consist of both the direction (i.e., the circular variable) and the
duration (i.e., the linear variable)(e.g., Rayner (2009)). Apart from
the social sciences data with a circular and linear outcome more
commonly occur in meteorology meteorology (García-Portugués, Crujeiras,
\& González-Manteiga, 2013), ecology (García-Portugués, Barros,
Crujeiras, González-Manteiga, \& Pereira, 2014) or marine research
(Lagona, Picone, Maruotti, \& Cosoli, 2015) \newline \indent Up until
now researchers studying cylindrical data had to rely on the linear
statistical methods to analyze their research results. Focusing on only
the circular variable or only on the linear variable. However, lately
more and more of these researchers acknowledge that linear methods are
not sufficient and call for new methods (Gurtman, 2011; Pennings, 2017b;
Wright, Pincus, Conroy, \& Hilsenroth, 2009) that take into account both
the circular and the linear component of these data. \newline
\indent The aim in the present paper is twofold. Firstly, we intend to
fill this gap in the literature on cylindrical models by adapting four
existing cylindrical models in such a way that they include a regression
of both the linear and circular component of a cylindrical variable onto
a set of covariates. From now on we will therefore refer to the
components of the cylindrical variable as outcome components. Secondly,
we will show how using these models can benefit the analysis of
circumplex data and cylindrical data in psychology in general. More
specifically we will show these benefits for interpersonal teacher data
from the field of educational psychology. Apart from modelling the
dependence between the linear and circular component of a cylindrical
variable we would also like to predict the two components from a set of
covariates in a regression model. In our analysis, we will show how this
can be done in four different ways using the adapted cylindrical models.

@@

Thus, data that consist of a variable and a circular variable is called
cylindrical data. A circular variable is different from the linear
variable in the sense that it is measured on a different scale. Figure
\ref{circline} shows the difference between a circular scale (right) and
a linear scale (left). The most important difference is that on a
circular scale the datapoints 0\(^\circ\) and 360\(^\circ\) are
connected and in fact represent the same number while on a linear scale
the two ends, \(-\infty\) and \(\infty\), are not connected and
consequently the values 0\(^\circ\) and 360\(^\circ\) are located on
different places on the scale. Both circular data and cylindrical data
require special analysis methods due to this periodicity in the scale of
a circular variable (see e.g.~Fisher (1995) for an introduction to
circular data and Mardia and Jupp (2000), Jammalamadaka and Sengupta
(2001) and Ley and Verdebout (2017) for a more elaborate
overview).\newline \indent In the literature, several methods have been
put forward to model the relation between the linear and circular
component of a cylindrical variable. Some of these are based on
regressing the linear component onto the circular component using the
following type of relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is
the linear component and \(\theta\) the circular component (Johnson \&
Wehrly, 1978; Mardia \& Sutton, 1978; Mastrantonio, Maruotti, \&
Jona-Lasinio, 2015). Others model the relation in a different way,
e.g.~by specifying a multivariate model for several linear and circular
variables and modelling their covariance matrix (Mastrantonio, 2018) or
by proposing a joint cylindrical distribution. For example, Abe and Ley
(2017) introduce a cylindrical distribution based on a Weibull
distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their
respective shape and concentration parameters. However, none of the
methods that have been proposed thus far include additional covariates
onto which both the circular and linear component are regressed.

@@

Communion (horizontal), see Figure \ref{QTI}. Different parts of this
space are characterized by different teacher behavior,
e.g.~\enquote{helpful} or \enquote{uncertain}. This two-dimensional
space is called the interpersonal circle/circumplex (IPC). The IPC is
``a continuous order with no beginning or end'' (Gurtman, 2009, p. 2).
We call such ordering a circumplex ordering and the IPC is therefore
often called the interpersonal circumplex. The ordering also implies
that scores on the IPC could be viewed as a circular variable. This
circular variable represents the type of interpersonal behavior that a
teacher shows towards his/her students.\newline \indent Cremers et al.
(2018a) explain the circular nature of the IPC data and analyze them as
such using a circular regression model. The two dimension scores, Agency
and Communion, can be converted to a circular score using the
two-argument arctangent function in
\eqref{PredVal}\footnote{The selection of

@@

scores; their respective 0 scores form the origin. Agency and Communion are
measured on a scale from 1 to 5 and for analysis purposes they are later
converted to scale ranging from -1 to 1. Scaling is however only considered an
issue in those instances where cylindrical data is derived from measurements in

@@

both dimensions separately providing us with new insights compared to a
separate analysis of the two dimensions that is standard in the
literature (see e.g.~Pennings et al. (2018), Wright et al. (2009) or

@@

The teacher data was collected between 2010 and 2015 and contains
several repeated measures on the IPC of 161 teachers. Measurements were
obtained using the QTI and taken in different years and classes. For
this paper we only consider one measurement, the first occasion (2010)
and largest class if data for multiple classes were available. This
results in a sample of 151 teachers. In addition to the type of
interpersonal behavior (IPC), the circular outcome, and the strength of
interpersonal behavior (IPC strength), the linear outcome, a teachers'
self-efficacy (\verb|SE|) concerning classroom management is used as
covariate in the analysis. In previous research, in psychology and
education it had been shown that higher self-efficacy is related to the
quality of interpersonal interactions (Locke \& Sadler, 2007;
\VANDER{Want}{Van der}{van der} Want et al., n.d.). After listwise
deletion of missings (\(3\) in total, only for the self-efficacy) we
have a sample of 148 teachers. Table \ref{Tableteacherdescriptives}
shows descriptives for the dataset. For the circular variable IPC we
show sample estimates for the circular mean \(\bar{\theta}\) and
concentration \(\hat{\rho}\). The circular concentration lies between 0,
meaning the data is not concentrated at all \emph{i.e.} spread over the
entire circle, and 1, meaning all data is concentrated at a single point
on the circle The population values of these parameters are usually, and
also in this paper, referred to as \(\mu\) (circular location) and
\(\kappa\) (circular concentration). For the linear variables (strength
IPC and SE) we show sample estimates of the linear mean and standard
deviation. Figure \ref{dataplot} is a scatterplot showing the relation
between the linear and circular outcome of the teacher data for teachers
with low SE (below 1 sd below the mean), average SE (between 1 sd below
and 1 sd above the mean) and high SE (above 1 sd above the mean).

@@

One of the goals of this paper is to show the benefits of cylindrical
methods for the analysis of circumplex data and cylindrical data in
psychology in general. To do so we decide to focus on four cylindrical
models. The models were selected for their relatively low complexity and
the ease with which a regression structure could be incorporated. But
also because they show different ways of modelling the the linear and
circular outcome and thereby illustrate a wider range of cylindrical
models available in the literature. All four cylindrical models contain
a set of \(q\) predictors \(\boldsymbol{x} = x_1, \dots, x_q\) and \(p\)
predictors \(\boldsymbol{z} = z_1, \dots, z_q\) for the linear and
circular outcomes, \(Y\) and \(\Theta\), respectively. The first two
models are based on a construction by Mastrantonio et al. (2015), while
the other models are extensions of the models from Abe and Ley (2017)
and Mastrantonio (2018). The four cylindrical models are introduced
separately in the subsections below. However, to provide a more succinct
overview and comparison of the four models, \ref{TableModels} gives an
overview of the similarities and differences between the models.

@@

data from the circumplex. Furthermore, in addition to being able to
assess the influence of covariates, the cylindrical models also provide
information about the dependence between the type and strength of
interpersonal behavior. We found that stronger behavior is associated
with those types of behavior that show higher scores on the Communion
dimension (and also on the Agency dimension). This translates to
stronger behavior being shown by teachers whose type of interpersonal
behavior ranges between 0\(^\circ\) and 90\(^\circ\), the
\enquote{helpful} and \enquote{directing} subtypes according to Figure
\ref{QTI}.
\textcolor{red}{Is dit nieuw/zijn er papers die hier iets over
zeggen?}\newline \indent As mentioned in the introducion data from the
interpersonal circumplex is not the only type of cylindrical data that
occurs in psychology. The methods presented in this paper are also of
use for research on human navigation and eye-tracking research.
Furthermore, even though cylindrical models are already used in fields
outside of psychology, the addition of a regression structure to the
models is of use in these fields as well. \newline \indent In terms of
interpretability, the CL-PN and Abe-Ley models perform best out of the
four cylindrical models. In the CL-GPN and GPN-SSN models the
interpretation of the parameters of the circular outcome component is
not straightforward, if at all possible. This is caused by the fact that
in addition to the mean vector the covariance matrix of the GPN
distribution affects the location of the circular data, making it
difficult to compute regression coefficients on the circle. Wang and
Gelfand (2013) state that Monte Carlo integration can be used to compute
a circular mean and variance for the GPN distribution. In future
research, this solution might be applied to the methods of Cremers et
al. (2018b) in order to compute circular coefficients for GPN
models.\newline \indent In terms of flexibility the GPN-SSN model scores
best. Multiple linear and circular outcomes can be included and we can
thus apply the model to multivariate cylindrical data. In addition the
GPN-SSN, the CL-GPN and CL-PN models are extendable to a mixed-effects
structure and can thus also be fit to longitudinal data (see
Nuñez-Antonio and Gutiérrez-Peña (2014) and Hernandez-Stumpfhauser et
al. (2016) for hierarchical/mixed-effects models for the PN and GPN
distributions respectively). For the Abe-Ley model this may also be
possible but has not been done in previous research for the conditional
distribution of its circular outcome (sine-skewed von Mises). Concerning
asymmetry, both the GPN-SSN as well as the Abe-Ley model allow for
non-symmetrical shapes of the distributions of both the linear and
circular outcome, while the CL-GPN model permits an asymmetric circular
outcome.\newline \indent The four cylindrical models that were modified
to the regression context in this paper are not the only cylindrical
distributions available from the literature. Other interesting
cylindrical distributions have been introduced by Fernández-Durán
(2007), Kato and Shimizu (2008) and Sugasawa (2015) (for more references
we refer to Chapter 2 of Ley and Verdebout (2017)). In the present study
we have decided not to include these distributions for reasons of space,
complexity of the models and ease of implementing a regression
structure. In future research however it would be interesting to
investigate other types of cylindrical distributions as well in order to
compare the interpretability, flexibility and model fit to the models
developed in the present study.\newline

@@

\leavevmode\hypertarget{ref-want2018selfefficacy}{}%
\VANDER{Want}{Van der}{van der} Want, A. C., Den Brok, P., Beijaard, D.,
Brekelmans, M., Claessens, L. C., \& Pennings, H. J. M. (n.d.). The
relation between teachers' interpersonal role identity and their
self-efficacy, burnout and work engagement. \emph{Professional
Development in Education}.
doi:\href{https://doi.org/10.1080/19415257.2018.1511453}{10.1080/19415257.2018.1511453}

>>>>

\authornote{

Correspondence concerning this article should be addressed to Jolien
Cremers, . E-mail:
\href{mailto:joliencremers@gmail.com}{\nolinkurl{joliencremers@gmail.com}}}

@@

In social sciences the use of cylindrical data is very common. Such data
consist of a linear and a circular component. Gurtman (2011) refers to
such data as vectors, with a directional measure (i.e., the circular
component) and a measure indicating the magnitude (i.e., linear
component). Many established models in psychology are often referred to
as circular or circumplex models, but those models are cylindrical.
Examples of such cylindrical models are the interpersonal
circle/circumplex (Leary, 1957; Wiggins, 1996; Wubbels, Brekelmans,
Brok, \& Tartwijk, 2006), the circumplex of affect (Russell, 1980), the
circumplex of human emotion (Plutchik, 1997) or the model of human
values (Schwartz, 1992). \newline \indent Also, many of the more recent
types of data that are studied in psychology are cylindrical. For
example, research on human navigation uses data where distance (i.e.,
linear component) and direction (i.e., the linear component) are of
interest (Chrastil \& Warren, 2017) or in eye-tracking, the saccade data
also consist of both the direction (i.e., the circular variable) and the
duration (i.e., the linear variable)(e.g., Rayner (2009)). Apart from
the social sciences data with a circular and linear outcome more
commonly occur in meteorology (García-Portugués, Crujeiras, \&
González-Manteiga, 2013), ecology (García-Portugués, Barros, Crujeiras,
González-Manteiga, \& Pereira, 2014) or marine research (Lagona, Picone,
Maruotti, \& Cosoli, 2015) \newline \indent Up until now researchers
studying cylindrical data had to rely on linear statistical methods to
analyze their research results. However, lately more and more of these
researchers acknowledge that linear methods are not sufficient and call
for new methods (Gurtman, 2011; Pennings, 2017b; Wright, Pincus, Conroy,
\& Hilsenroth, 2009) that take into account both the circular and the
linear component of these data. \newline \indent The aim in the present
paper is twofold. Firstly, we we intend to fill the above mentioned gap
in the literature by showing the use of cylindrical models can benefit
the analysis of circumplex data and cylindrical data in psychology in
general. More specifically we will show these benefits for interpersonal
teacher data from the field of educational psychology. Apart from
modelling the dependence between the linear and circular component of a
cylindrical variable we would also like to predict the two components
from a set of covariates in a regression model. Our second aim therefore
is to adapt several existing cylindrical models in such a way that they
include a regression of both the linear and circular component of a
cylindrical variable onto a set of covariates. From now on we will
therefore refer to the components of the cylindrical variable as outcome
components. These adapted circular models are then used to analyse the
teacher data.

@@

Data that consist of a variable and a circular variable is called
cylindrical data. A circular variable is different from the linear
variable in the sense that it is measured on a different scale. Figure
\ref{circline} shows the difference between a circular scale (right) and
a linear scale (left). The most important difference is that on a
circular scale the datapoints 0\(^\circ\) and 360\(^\circ\) are
connected and in fact represent the same number while on a linear scale
the two ends, \(-\infty\) and \(\infty\), are not connected and
consequently the values 0\(^\circ\) and 360\(^\circ\) are located on
different places on the scale. Both circular data and cylindrical data
require special analysis methods due to this periodicity in the scale of
a circular variable (see e.g.~Fisher (1995) for an introduction to
circular data and Mardia and Jupp (2000), Jammalamadaka and Sengupta
(2001) and Ley and Verdebout (2017) for a more elaborate
overview).\newline \indent In the literature, several methods have been
put forward to model the relation between the linear and circular
component of a cylindrical variable. Some of these are based on
regressing the linear component onto the circular component using the
following type of relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is
the linear component and \(\theta\) the circular component (Johnson \&
Wehrly, 1978; Mardia \& Sutton, 1978; Mastrantonio, Maruotti, \&
Jona-Lasinio, 2015). Others model the relation in a different way,
e.g.~by specifying a multivariate model for several linear and circular
variables and modelling their covariance matrix (Mastrantonio, 2018) or
by proposing a joint cylindrical distribution. For example, Abe and Ley
(2017) introduce a cylindrical distribution based on a Weibull
distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their
respective shape and concentration parameters. However, none of the
methods that have been proposed thus far include additional covariates
onto which both the circular and linear component are regressed.

@@

Communion (horizontal), see Figure \ref{QTI}. This space is called the
interpersonal circle/circumplex (IPC) and different parts of this space
are characterized by different teacher behavior, e.g.~\enquote{helpful}
or \enquote{uncertain}. The IPC is ``a continuous order with no
beginning or end'' (Gurtman, 2009, p. 2). We call such ordering a
circumplex ordering and the IPC is therefore often called the
interpersonal circumplex. The ordering also implies that scores on the
IPC could be viewed as a circular variable. This circular variable
represents the type of interpersonal behavior that a teacher shows
towards his/her students.\newline \indent Cremers et al. (2018a) explain
the circular nature of the IPC data and analyze them as such using a
circular regression model. The two dimension scores, Agency and
Communion, can be converted to a circular score using the two-argument
arctangent function in \eqref{PredVal}\footnote{The selection of

@@

scores. Agency and Communion are measured on a scale from 1 to 5 and for
analysis purposes they are later converted to scale ranging from -1 to 1.  Their
respective 0 scores form the origin. Scaling is however only considered an issue
in those instances where cylindrical data is derived from measurements in

@@

both dimensions separately. This provides us with new insights compared
to a separate analysis of the two dimensions that is standard in the
literature (see e.g. Pennings et al. (2018), Wright et al. (2009) or

@@

The teacher data was collected between 2010 and 2015 and contains
several repeated measures on the IPC of 161 teachers. Measurements were
obtained using the QTI and taken in different years and classes. For
this paper we only consider one measurement, the first occasion (2010)
and largest class if data for multiple classes were available. This
results in a sample of 151 teachers. In addition to the type of
interpersonal behavior (IPC), the circular outcome, and the strength of
interpersonal behavior (IPC strength), the linear outcome, a teachers'
self-efficacy (\verb|SE|) concerning classroom management is used as
covariate in the analysis. In previous research, in psychology and
education it has been shown that higher self-efficacy is related to the
quality of interpersonal interactions (Locke \& Sadler, 2007;
\VANDER{Want}{Van der}{van der} Want et al., 2018). After listwise
deletion of missings (\(3\) in total, only for the self-efficacy) we
have a sample of 148 teachers. Table \ref{Tableteacherdescriptives}
shows descriptives for the dataset. For the circular variable IPC we
show sample estimates for the circular mean \(\bar{\theta}\) and
concentration \(\hat{\rho}\). The circular concentration lies between 0,
meaning the data is not concentrated at all \emph{i.e.} spread over the
entire circle, and 1, meaning all data is concentrated at a single point
on the circle The population values of these parameters are usually, and
also in this paper, referred to as \(\mu\) (circular location) and
\(\kappa\) (circular concentration). For the linear variables (strength
IPC and SE) we show sample estimates of the linear mean and standard
deviation (sd). Figure \ref{dataplot} is a scatterplot showing the
relation between the linear and circular outcome of the teacher data for
teachers with low SE (below 1 sd below the mean), average SE (between 1
sd below and 1 sd above the mean) and high SE (above 1 sd above the
mean).

@@

One of the goals of this paper is to show the benefits of cylindrical
methods for the analysis of circumplex data and cylindrical data in
psychology in general. To do so we decide to focus on four cylindrical
models. The models were selected for their relatively low complexity and
the ease with which a regression structure could be incorporated. But
also because they show different ways of modelling the the linear and
circular outcome and thereby illustrate a wider range of cylindrical
models available in the literature. The cylindrical models contain a set
of \(q\) predictors \(\boldsymbol{x} = x_1, \dots, x_q\) and \(p\)
predictors \(\boldsymbol{z} = z_1, \dots, z_q\) for the linear and
circular outcomes, \(Y\) and \(\Theta\), respectively. The first two
models are based on a construction by Mastrantonio et al. (2015), while
the other models are extensions of the models from Abe and Ley (2017)
and Mastrantonio (2018). The four cylindrical models are introduced
separately in the subsections below. However, to provide a more succinct
overview and comparison of the four models, Table \ref{TableModels}
gives an overview of the similarities and differences between the
models.

@@

data from the circumplex. \newline \indent Furthermore, in addition to
being able to assess the influence of covariates, the cylindrical models
also provide information about the dependence between the type and
strength of interpersonal behavior. We found that stronger behavior is
associated with higher scores on the Communion and in some models also
the Agency dimension. This implies that teachers whose type of
interpersonal behavior ranges between 0\(^\circ\) and 90\(^\circ\), the
\enquote{helpful} and \enquote{directing} subtypes are stronger in their
behavior than teachers of the other subtypes.\newline \indent As
mentioned in the introduction, data from the interpersonal circumplex is
not the only type of cylindrical data that occurs in psychology. The
methods presented in this paper are also of use for research on human
navigation and eye-tracking research. Furthermore, even though
cylindrical models are already used in fields outside of psychology, the
addition of a regression structure to the models is of use in these
fields as well. \newline \indent In terms of interpretability, the CL-PN
and Abe-Ley models perform best out of the four cylindrical models. In
the CL-GPN and GPN-SSN models the interpretation of the parameters of
the circular outcome component is not straightforward, if at all
possible. This is caused by the fact that in addition to the mean vector
the covariance matrix of the GPN distribution affects the location of
the circular data, making it difficult to compute regression
coefficients on the circle. Wang and Gelfand (2013) state that Monte
Carlo integration can be used to compute a circular mean and variance
for the GPN distribution. In future research, this solution might be
applied to the methods of Cremers et al. (2018b) in order to compute
circular coefficients for GPN models.\newline \indent In terms of
flexibility the GPN-SSN model scores best. Multiple linear and circular
outcomes can be included and we can thus apply the model to multivariate
cylindrical data. In addition the GPN-SSN, the CL-GPN and CL-PN models
are extendable to a mixed-effects structure and can thus also be fit to
longitudinal data (see Nuñez-Antonio and Gutiérrez-Peña (2014) and
Hernandez-Stumpfhauser et al. (2016) for hierarchical/mixed-effects
models for the PN and GPN distributions respectively). For the Abe-Ley
model this may also be possible but has not been done in previous
research for the conditional distribution of its circular outcome
(sine-skewed von Mises). Concerning asymmetry, both the GPN-SSN as well
as the Abe-Ley model allow for non-symmetrical shapes of the
distributions of both the linear and circular outcome, while the CL-GPN
model permits an asymmetric circular outcome.\newline \indent The four
cylindrical models that were modified to the regression context in this
paper are not the only cylindrical distributions available from the
literature. Other interesting cylindrical distributions have been
introduced by Fernández-Durán (2007), Kato and Shimizu (2008) and
Sugasawa (2015) (for more references we refer to Chapter 2 of Ley and
Verdebout (2017)). In the present study we have decided not to include
these distributions for reasons of space, complexity of the models and
ease of implementing a regression structure. In future research however
it would be interesting to investigate other types of cylindrical
distributions as well in order to compare the interpretability,
flexibility and model fit to the models developed in the present
study.\newline

@@

\leavevmode\hypertarget{ref-want2018selfefficacy}{}%
\VANDER{Want}{Van der}{van der} Want, A. C., Den Brok, P., Beijaard, D.,
Brekelmans, M., Claessens, L. C., \& Pennings, H. J. M. (2018). The
relation between teachers' interpersonal role identity and their
self-efficacy, burnout and work engagement. \emph{Professional
Development in Education}, Advance Online Publication.
doi:\href{https://doi.org/10.1080/19415257.2018.1511453}{10.1080/19415257.2018.1511453}
